---
title: 'assignment2: topic model'
author: "yue shen"
date: "2026-09-23"
output:
  html_document:
    code_folding: hide
  pdf_document: default
  word_document: default
---
# Research Question
My original research question asks whether journalistic, institutional, and corporate sources differ in how they frame the benefits and risks of artificial intelligence (AI) in special education and in the types of evidence used to support these claims.

However, during the initial exploration of the corpus, I found that the retrieved documents covered a broader range of AI and education topics than expected, while relatively fewer documents focused specifically on AI in special education. Therefore, for this preliminary topic-modeling analysis, I use a broader exploratory question:
**What major topics emerge from the retrieved corpus on AI and special education, and to what extent do these topics represent substantively relevant educational and disability-related discourse?**

The original comparison of source type, framing pattern, and evidence type will remain relevant for later stages of the project after the corpus and topic structure are better understood.






```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(
  root.dir = "/Users/koyohoma/Documents/UTD/2026 Fall/EPPS6355/EPPS6355"
)
```

```{r}
library(quanteda)
library(readtext)
library(stm)
```
# Text Data

The text data were collected from Nexis Uni through the UTD Library. The search strategy included two concept blocks: AI-related terms and special-education-related terms. To improve topical relevance, the two concept blocks were connected using the Nexis Uni proximity operator `W/P`, requiring the terms to occur within the same paragraph.

The search included AI-related terms such as "artificial intelligence," "machine learning," "generative AI," "large language model," ChatGPT, Claude AI, Google Gemini, and Microsoft Copilot, together with special-education-related terms such as "special education," "IEP," "special needs education," and "individualized education program."

For Assignment 2, I downloaded a subset of the search results in two batches. The downloaded data initially contained 974 RTF files.

## Data Import and Preprocessing
The downloaded RTF files were imported into R and converted into a document-level text dataset. Two downloaded files were Nexis-generated document lists rather than news documents and were removed, leaving 972 documents for the initial analysis.

The text was then converted into a `quanteda` corpus and tokenized. Text was converted to lowercase, while punctuation and symbols were removed. Standard English stopwords were removed, although several negation and modal terms such as *not*, *can*, *could*, *may*, *might*, *should*, and *must* were retained because they may carry meaningful information about how AI is discussed.

Several multiword expressions, including *artificial intelligence* and *special education*, were compounded so that they could be treated as single features. The processed tokens were then converted into a document-feature matrix (DFM).

```{r}

# Read all real RTF files from both downloaded batches (Chatgpt)
rtf_files <- c(
  list.files(
    "raw data/Files1-500",
    pattern = "\\.rtf$",
    full.names = TRUE,
    ignore.case = TRUE
  ),
  list.files(
    "raw data/Files 501-1000",
    pattern = "\\.rtf$",
    full.names = TRUE,
    ignore.case = TRUE
  )
)

length(rtf_files)

sum(rtf_files == "")

sum(is.na(rtf_files))

sum(!file.exists(rtf_files))
library(striprtf)

ML1000 <- data.frame(
  doc_id = rtf_files,
  text = sapply(
    rtf_files,
    function(f) paste(read_rtf(f), collapse = "\n")
  ),
  stringsAsFactors = FALSE
)

nrow(ML1000)

```
# Remove Nexis document-list files (ChatGpt)
```{r}
ML1000 <- ML1000[
  !grepl("doclist", ML1000$doc_id, ignore.case = TRUE),
]

nrow(ML1000)
```

# Reuse preprocessing decisions from Assignment 1
```{r}
my_corpus <- corpus(
  ML_clean$text,
  docnames = ML_clean$doc_id
)

my_tokens <- tokens(
  my_corpus,
  remove_punct = TRUE,
  remove_symbols = TRUE,
  remove_numbers = TRUE
) |>
  tokens_tolower()

my_tokens <- tokens_compound(
  my_tokens,
  pattern = phrase(c(
    "artificial intelligence",
    "special education",
    "individualized education program",
    "individualized education plan"
  ))
)


my_stopwords <- setdiff(
  stopwords("en"),
  c("not", "no", "nor",
    "can", "could", "may", "might",
    "should", "must")
)

my_tokens <- tokens_remove(
  my_tokens,
  pattern = my_stopwords
)


my_dfm <- dfm(my_tokens)

```
# To check the DFM information
```{r}
ndoc(my_dfm)
nfeat(my_dfm)

topfeatures(my_dfm, 30)
```
# To remove something irrelevant 
```{r}
extra_stopwords <- c(
  "said",
  "also",
  "new",
  "news",
  "year",
  "end",
  "words"
)

my_tokens <- tokens_remove(
  my_tokens,
  pattern = extra_stopwords
)

my_dfm <- dfm(my_tokens)

topfeatures(my_dfm, 30)

my_dfm_trimmed <- dfm_trim(
  my_dfm,
  min_docfreq = 3 #a feature at least should appear in 3 articles to be kept
)
```
# adapt DFM to STM format (Chatgpt)
```{r}

library(stm)
stm_input_trimmed <- convert(my_dfm_trimmed, to = "stm") #convert dfm to stm 
```

# build a five-topic model
```{r}
set.seed (6261705)

my_stm_trimmed <- stm(
  stm_input_trimmed$documents,
  stm_input_trimmed$vocab,
  K = 5,
  init.type = "Spectral"
)

labelTopics(my_stm_trimmed, n = 10)
```

```{r}
findThoughts(
  my_stm_trimmed,
  texts = ML_clean$text,
  topics = c(1, 2, 3, 4, 5),
  n = 2
)
```


# Preliminary Corpus Exploration
An initial five-topic STM was fitted to the 972-document corpus. Inspection of the most probable topic words and representative documents revealed substantial noise in the corpus. Several topics were dominated by unrelated medical, financial, and technology-industry content rather than AI in educational contexts.

Further inspection showed that some Nexis files were composite news digests rather than single coherent articles. In particular, sources such as *News Bites* and *Global Round Up* frequently combined multiple unrelated news items within one downloaded document. Because the intended unit of analysis in this project is one coherent article, these composite documents were excluded.

After removing these sources, the working corpus was reduced to **670 documents**. The STM will be re-estimated using this refined corpus, and topic interpretations from the preliminary model are therefore treated as diagnostic rather than final results.

#Finding and Limitation
Validation using representative documents showed that the five-topic model captured both substantively relevant and irrelevant content. Topics 4 and 5 represented meaningful educational and disability-related discourse, including classroom uses of AI and AI/ML applications in autism. In contrast, Topics 2 and 3 revealed systematic retrieval noise. In particular, the acronym “IEP” sometimes referred to terms unrelated to Individualized Education Programs, such as the Institute for Economics and Peace, while similar character strings also appeared in technology product names. Topic 1 represented a more peripheral form of relevance, focusing primarily on AI in academic research while mentioning special education only incidentally. These findings indicate that keyword-based retrieval requires further validation before the original source-type and framing comparisons can be conducted



```

# Topic 1 validation
```{r}
findThoughts(
  my_stm_trimmed,
  texts = ML1000$text,
  topics = 1,
  n = 3
)
```

# create filter (ChatGpt)
```{r}

ai_terms <- paste(
  c("artificial intelligence", "\\bAI\\b"), #word boundary, find AI instead of some word contains "ai"
  collapse = "|"
)

special_ed_terms <- paste(
  c(
    "special education",
    "individualized education program",
    "individualized education plan",
    "students with disabilities",
    "children with disabilities"
  ),
  collapse = "|"
)

disability_terms <- paste(
  c(
    "autism",
    "\\bASD\\b",
    "disability",
    "disabilities"
  ),
  collapse = "|"
)

education_terms <- paste(
  c(
    "student",
    "students",
    "school",
    "schools",
    "teacher",
    "teachers",
    "classroom",
    "education",
    "learning"
  ),
  collapse = "|"
)

ML1000$relevant <- 
  grepl(ai_terms, ML1000$text, ignore.case = TRUE) &
  (
    grepl(special_ed_terms, ML1000$text, ignore.case = TRUE) |
    (
      grepl(disability_terms, ML1000$text, ignore.case = TRUE) & #disability term should appear in educational context
      grepl(education_terms, ML1000$text, ignore.case = TRUE)
    )
  )

table(ML1000$relevant)
```

```{r}
# 🔧 Validation of the relevance filter

set.seed(123)

true_ids <- sample(
  which(ML1000$relevant == TRUE),
  5
)

false_ids <- sample(
  which(ML1000$relevant == FALSE),
  5
)

substr(ML1000$text[true_ids], 1, 800)
substr(ML1000$text[false_ids], 1, 800)
```

```{r}
# 🔧 Check why each article passed the relevance filter (Chatgpt)

ML1000$hit_ai <- grepl(
  ai_terms,
  ML1000$text,
  ignore.case = TRUE
)

ML1000$hit_special_ed <- grepl(
  special_ed_terms,
  ML1000$text,
  ignore.case = TRUE
)

ML1000$hit_disability <- grepl(
  disability_terms,
  ML1000$text,
  ignore.case = TRUE
)

ML1000$hit_education <- grepl(
  education_terms,
  ML1000$text,
  ignore.case = TRUE
)

ML1000[
  true_ids,
  c(
    "doc_id",
    "hit_ai",
    "hit_special_ed",
    "hit_disability",
    "hit_education"
  )
]
```

```{r}
# 🔧 Show the text around the matched special-education term (Chatgpt)

show_context <- function(text, pattern, window = 250) {
  
  clean_text <- gsub("\\s+", " ", text)
  
  match <- regexpr(
    pattern,
    clean_text,
    ignore.case = TRUE,
    perl = TRUE
  )
  
  if (match[1] == -1) {
    return("No match")
  }
  
  start <- max(1, match[1] - window)
  end <- min(
    nchar(clean_text),
    match[1] + attr(match, "match.length") + window
  )
  
  substr(clean_text, start, end)
}

sapply(
  ML1000$text[true_ids],
  show_context,
  pattern = special_ed_terms
)
```

```{r}
# 🔧 Diagnose composite news-digest documents (Chatgpt)

ML1000$is_news_bites <- grepl(
  "News Bites",
  substr(ML1000$text, 1, 300),
  ignore.case = TRUE
)

ML1000$is_global_roundup <- grepl(
  "Global Round Up",
  substr(ML1000$text, 1, 300),
  ignore.case = TRUE
)

table(ML1000$is_news_bites)
table(ML1000$is_global_roundup)

sum(
  ML1000$is_news_bites |
  ML1000$is_global_roundup
)
```
```{r}

# randomly see 5 articles
set.seed(123)

digest_ids <- which(
  ML1000$is_news_bites |
  ML1000$is_global_roundup
)

sample_digest <- sample(digest_ids, 5)

ML1000$doc_id[sample_digest]
```
# exclude all the articles from News Bite and Global Roundup
```{r}
ML_clean <- ML1000 [!(ML1000$is_news_bites|ML1000$is_global_roundup),]
nrow(ML_clean)


```



